---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: true
---

```{r setoptions, echo=FALSE}
options(scipen=999)
```

## Loading and preprocessing the data

##### Reading the data into the 'activitydata' variable and showing the dimensions.

```{r echo=TRUE}
activityData <- read.csv(unz("activity.zip", filename = "activity.csv"))
str(activityData)
```

##### Processing data to give proper date format
```{r echo=TRUE}
activityData$date <- as.Date(activityData$date)
str(activityData)
```

## What is mean total number of steps taken per day?

##### Plotting the histogram of total steps taken each day
```{r echo=TRUE}
totStepsEachDay <- aggregate(x = activityData$steps, by = list(activityData$date), FUN = sum, na.rm = TRUE)
names(totStepsEachDay) <- c("Date", "Total.Steps")
hist(x = totStepsEachDay$Total.Steps, main = "Histogram of Total steps each day", xlab = "Total steps each day", col = "chocolate1")

meanSteps <- mean(totStepsEachDay$Total.Steps)
medianSteps <- median(totStepsEachDay$Total.Steps)
```

###### Mean of the total number of steps taken each day: `r meanSteps`
###### Median of the total number of steps taken each day: `r medianSteps`

## What is the average daily activity pattern?
##### Calculating the average number of steps taken based on the 5 min interval, across all days
```{r echo=TRUE}
avgStepsInInterval <- aggregate(x = activityData$steps, by = list(activityData$interval), FUN = mean, na.rm = TRUE)
names(avgStepsInInterval) <- c("Interval", "Average.Steps")
```

##### Plotting Activity Daily Pattern
```{r echo=TRUE}
plot(x = avgStepsInInterval$Interval, y = avgStepsInInterval$Average.Steps, type = "l", xlab = "5-minute Interval", ylab = "Average number of steps", main = "Average Daily Activity Pattern", col = "chocolate1")
```

##### Finding the 5-minute interval with maximum average steps taken across all days
```{r echo=TRUE}
reqdInterval <- avgStepsInInterval[avgStepsInInterval$Average.Steps == max(avgStepsInInterval$Average.Steps),]$Interval
```

###### The 5-minute interval, on average across all the days in the dataset, with the maximum number of steps: `r reqdInterval` - `r reqdInterval+5`

## Imputing missing values

##### Calculating the total number of missing values in the dataset
```{r echo=TRUE}
countNA <- sum(!complete.cases(activityData))
```

Number of rows with missing values: `r countNA`
 
##### Filling missing values with the mean for the respective 5-minute interval
```{r echo=TRUE}
activityDataSansNA <- activityData
for (i in 1:nrow(activityDataSansNA)){
    if(is.na(activityDataSansNA[i, "steps"])){
        activityDataSansNA[i, "steps"] = 
            ceiling(avgStepsInInterval[avgStepsInInterval$Interval ==
                activityDataSansNA[i,"interval"],]$Average.Steps)
    }
}
```

Printing dimensions of the new dataset
```{r echo=TRUE}
str(activityDataSansNA)
```

##### Plotting the histogram of total number of steps taken each day after filling NA's
```{r echo=TRUE}
totStepsEachDaySansNA <- aggregate(x = activityDataSansNA$steps, by = list(activityDataSansNA$date), FUN = sum, na.rm = TRUE)
names(totStepsEachDaySansNA) <- c("Date", "Total.Steps")
hist(x = totStepsEachDaySansNA$Total.Steps, main = "Histogram of Total steps each day", xlab = "Total steps each day", col = "chocolate1")
meanSansNA <- mean(totStepsEachDaySansNA$Total.Steps)
medianSansNA <- median(totStepsEachDaySansNA$Total.Steps)
```

###### New mean of the total number of steps taken each day: `r meanSansNA`
###### New median of the total number of steps taken each day: `r medianSansNA`

###### Mean and median values have increased after filling the missing values.

```{r echo=TRUE}
totStepsAllDays <- sum(totStepsEachDay$Total.Steps, na.rm = TRUE)
newTotStepsAllDays <- sum(totStepsEachDaySansNA$Total.Steps)
```
###### Increase in the total number of steps after imputing: `r newTotStepsAllDays - totStepsAllDays`

## Are there differences in activity patterns between weekdays and weekends?

###### Adding variables to identify weekday and weekends
```{r echo=TRUE}
activityDataSansNA$day <- weekdays(activityDataSansNA$date,abbreviate = TRUE)
activityDataSansNA$daytype[(activityDataSansNA$day == "Sun" | activityDataSansNA$day == "Sat")] <- "weekend"
activityDataSansNA$daytype[!(activityDataSansNA$day == "Sun" | activityDataSansNA$day == "Sat")] <- "weekday"
activityDataSansNA$daytype <- as.factor(activityDataSansNA$daytype)
```

Printing dimensions of the new dataset
```{r echo=TRUE}
str(activityDataSansNA)
```

##### Plotting Activity Pattern across Weekends and Weekdays

```{r echo=TRUE}
```{r echo=TRUE}
avgStepsInIntervalDayType <- aggregate(steps ~ interval+daytype, activityDataSansNA, mean)
library(ggplot2)
qplot(interval, steps, data = avgStepsInIntervalDayType, geom = c("line"), xlab = "Interval", ylab = "Average No of Steps", col = "chocolate1", main="") + facet_wrap(~daytype, ncol = 1) + guides(colour=FALSE)

```
